---
title: "When Are Homicides Most Common?"
author: "Tommy Gavagan"
theme: superhero
fontsize: 16px
engine: knitr
project:
  type: website
  output-dir: docs
format:
  closeread-html:
    cr-section:
      layout: sidebar-left
    cr-style:
      section-background-color: "#f3f6fa"
      narrative-background-color-overlay: "#ffffff"
      narrative-text-color-overlay: "#1f1f1f"
      narrative-border-radius: 8px
      narrative-overlay-max-width: 58%
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(plotly)
library(htmltools)

homicides <- read.csv("homicide-data.csv", fileEncoding = "latin1")

month_order <- c(
  "January","February","March","April","May","June",
  "July","August","September","October","November","December"
)

homicides_clean <- homicides |>
  mutate(
    reported_date = as.Date(as.character(reported_date), "%Y%m%d"),
    month = factor(format(reported_date, "%B"), levels = month_order),
    city = trimws(city)
  ) |>
  filter(!is.na(month), city != "")

all_months <- tibble(month = factor(month_order, levels = month_order))
all_cities <- sort(unique(homicides_clean$city))

monthly_all <- homicides_clean |>
  count(month, name = "homicides") |>
  right_join(all_months, by = "month") |>
  mutate(homicides = replace_na(homicides, 0)) |>
  arrange(month)

city_monthly <- homicides_clean |>
  count(city, month, name = "homicides") |>
  right_join(
    expand_grid(city = all_cities, month = factor(month_order, levels = month_order)),
    by = c("city","month")
  ) |>
  mutate(homicides = replace_na(homicides, 0)) |>
  arrange(city, month)

peak_month <- monthly_all |> slice_max(homicides, n = 1)
low_month  <- monthly_all |> slice_min(homicides, n = 1)

p <- plot_ly()

p <- p |>
  add_bars(
    data = monthly_all,
    x = ~month,
    y = ~homicides,
    name = "All Cities",
    visible = TRUE,
    hovertemplate = "<b>Month:</b> %{x}<br><b>Homicides:</b> %{y}<extra></extra>"
  )

for (i in seq_along(all_cities)) {
  city_data <- city_monthly |> filter(city == all_cities[i])

  p <- p |>
    add_bars(
      data = city_data,
      x = ~month,
      y = ~homicides,
      name = all_cities[i],
      visible = FALSE,
      hovertemplate = paste(
        "<b>City:</b>", all_cities[i], "<br>",
        "<b>Month:</b> %{x}<br>",
        "<b>Homicides:</b> %{y}<extra></extra>"
      )
    )
}

buttons <- list()

buttons[[1]] <- list(
  method = "update",
  args = list(list(visible = c(TRUE, rep(FALSE, length(all_cities))))),
  label = "All Cities"
)

for (i in seq_along(all_cities)) {
  visible <- rep(FALSE, length(all_cities)+1)
  visible[i+1] <- TRUE

  buttons[[i+1]] <- list(
    method = "update",
    args = list(list(visible = visible)),
    label = all_cities[i]
  )
}

homicide_plot <- p |>
  layout(
    title = "Homicides by Month",
    xaxis = list(title = "Month"),
    yaxis = list(title = "Count"),
    updatemenus = list(list(type = "dropdown", buttons = buttons)),
    margin = list(r = 160)
  )
```

# Introduction

Homicide is often discussed using yearly totals, but those numbers hide patterns 
within the year. Instead of asking only where violence happens, this project 
asks a simpler question: **when does it happen most often?**

:::::::: cr-section

## Why this matters

If homicides consistently peak in certain months, that suggests the pattern is not random.
Understanding timing can help show more broad behavioral/environmental influences.

## Dataset overview

This dataset includes reported homicide cases across multiple U.S. cities,
each with a reported date. By grouping cases by month, 
we can compare how counts change across the calendar year.
(https://www.kaggle.com/datasets/joebeachcapital/homicides/data)

## The question

**Do homicides follow a seasonal pattern? And does that pattern change by city?**

## Explore the data

Use the dropdown to switch between all cities and individual cities. Hover to see exact counts.

@cr-plot

::: {#cr-plot}
```{r}
homicide_plot
```
:::

## What stands out

The distribution is not uniform. Some months clearly have higher counts than others.

- Highest month: **`r peak_month$month`**
- Lowest month: **`r low_month$month`**

## Why the city filter matters

The overall trend might hide differences across cities. 
Some cities may have sharper peaks, 
while others remain more consistent throughout the year.

Use the dropdown to test this.

## Open exploration

This graphic raises a couple follow-up questions:

- Are warmer months consistently higher?
- Do certain cities drive the pattern?
- Would results change if adjusted for population?

The goal is not to answer everything, but to let you explore patterns freely.

::::::::

# Conclusion

By focusing on monthly patterns instead of yearly totals, this visualization 
reveals a clearer and more interactive view of when homicides occur. 
The addition of city-level filtering allows users to move beyond averages
and explore variation across locations.
```